---
title: "aline-propensity-score"
author: "Alistair Johnson"
date: "May 15, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Analysis of arterial line dataset

This notebook creates a propensity score using a dataset of patients with indwelling arterial catheters (IACs). The propensity score is built using physiology and administrative data to predict the need for an IAC. Patients are then matched, and we statistically compare the mortality rate in the two matched groups.

## Load data

First, we load the data and convert some variables into factors. Note this code assumes that you have the csv file available in "~/git/mimic-code/notebooks/aline/".

```{r load}
wdpath = paste(path.expand("~"),'git/mimic-code/notebooks/aline/',sep='/')
setwd(wdpath)
dataset = read.csv(file="aline_data.csv",head=TRUE,sep=",")
```


```{r factorize, echo = FALSE}
dataset$icustay_id = factor(dataset$icustay_id)
dataset$day_28_flag = factor(dataset$day_28_flag, levels=c(0,1))
dataset$gender = factor(dataset$gender, levels=c("F","M"))
dataset$day_icu_intime = factor(dataset$day_icu_intime)
dataset$hour_icu_intime = factor(dataset$hour_icu_intime)
dataset$icu_hour_flag = factor(dataset$icu_hour_flag, levels=c(0,1))
#dataset$sepsis_flag = factor(dataset$sepsis_flag, levels=c(0,1))
dataset$sedative_flag = factor(dataset$sedative_flag, levels=c(0,1))
dataset$fentanyl_flag = factor(dataset$fentanyl_flag, levels=c(0,1))
dataset$midazolam_flag = factor(dataset$midazolam_flag, levels=c(0,1))
dataset$propofol_flag = factor(dataset$propofol_flag, levels=c(0,1))
#dataset$dilaudid_flag = factor(dataset$dilaudid_flag, levels=c(0,1))
dataset$chf_flag = factor(dataset$chf_flag, levels=c(0,1))
dataset$afib_flag = factor(dataset$afib_flag, levels=c(0,1))
dataset$renal_flag = factor(dataset$renal_flag, levels=c(0,1))
dataset$liver_flag = factor(dataset$liver_flag, levels=c(0,1))
dataset$copd_flag = factor(dataset$copd_flag, levels=c(0,1))
dataset$cad_flag = factor(dataset$cad_flag, levels=c(0,1))
dataset$stroke_flag = factor(dataset$stroke_flag, levels=c(0,1))
dataset$malignancy_flag = factor(dataset$malignancy_flag, levels=c(0,1))
dataset$respfail_flag = factor(dataset$respfail_flag, levels=c(0,1))
dataset$ards_flag = factor(dataset$ards_flag, levels=c(0,1))
dataset$pneumonia_flag = factor(dataset$pneumonia_flag, levels=c(0,1))

# custom factor
dataset$service_surg = factor( dataset$service_unit == 'SURG', levels=c(FALSE,TRUE))
```

```{r impute, echo = FALSE}
# we could impute data if we like - e.g. the below imputes the mean
# we currently do complete case analysis however
imputeFlag = 0
if (imputeFlag != 0){
  print("Imputing missing data for some features...")
for (col in c("weight_first","temp_first","spo2_first",
              "bun_first","creatinine_first", "chloride_first", "hgb_first",
              "platelet_first", "potassium_first", "sodium_first", "tco2_first", "wbc_first"))
{
  print(paste("Imputing data for: ", col))
  dataset[is.na(dataset[,col]),col] = mean(dataset[,col], na.rm=TRUE)
}
}
```


## Propensity score model

Now, we build a logistic regression to predict the need for an arterial line catheter from physiology and administrative data.

```{r propensity}
library(Matching)

set.seed(43770)

# fit GLM
glm_fitted = glm(aline_flag ~  age + gender + weight_first + 
                      # bmi +  
                     sofa_first + service_surg + day_icu_intime + hour_icu_intime + 
                     #icu_hour_flag +
                     # sedative usage
                     # sedative_flag + fentanyl_flag + midazolam_flag + propofol_flag +
                     # comorbidities
                     chf_flag + afib_flag + renal_flag + 
                     liver_flag + copd_flag + cad_flag +
                     stroke_flag + malignancy_flag + respfail_flag + 
                     # ards_flag + pneumonia_flag + 
                     # vitals
                     map_first + hr_first + temp_first + spo2_first + 
                     # labs
                     bun_first + chloride_first + creatinine_first + 
                     hgb_first + platelet_first + potassium_first +
                     sodium_first + tco2_first + wbc_first, data=dataset, na.action = na.exclude)

X <- fitted(glm_fitted)
y <- dataset$day_28_flag
Tr <- dataset$aline_flag

# remove the rows with missing data
idxMiss = is.na(X)
print(paste('Missing data for ', sum(idxMiss)))
X <- X[!idxMiss]
y <- y[!idxMiss]
Tr <- Tr[!idxMiss]
dat <- dataset[!idxMiss,]
```

The above shows how many rows we have excluded due to missing data (this analysis uses complete case analysis).

We have built the propensity score using logistic regression in the previous block.
We now use the `Matching` package to match patients with a caliper size of 0.01.
After matching, we'll apply McNemar's test for paired samples to determine if patients with and without an a-line have a difference in mortality.

```{r ps}
ps <- Match(Y=NULL, Tr=Tr, X=X, M=1, estimand='ATC', caliper=0.01, exact=FALSE, replace=FALSE);

# jesse's propensity score workshop -> get pairs with treatment/outcome as cols

outcome <- data.frame(aline_pt=dat[ps$index.treated,"hosp_exp_flag"], match_pt=dat[ps$index.control,"hosp_exp_flag"])
head(outcome)

# mcnemar's test to see if iac related to mort (test should use matched pairs)
tab.match1 <- table(outcome$aline_pt,outcome$match_pt,dnn=c("Aline","Matched Control"))
tab.match1
tab.match1[1,2]/tab.match1[2,1]
paste("95% Confint", round(exp(c(log(tab.match1[2,1]/tab.match1[1,2]) - qnorm(0.975)*sqrt(1/tab.match1[1,2] +1/tab.match1[2,1]),log(tab.match1[2,1]/tab.match1[1,2]) + qnorm(0.975)*sqrt(1/tab.match1[1,2] +1/tab.match1[2,1])) ),2))
mcnemar.test(tab.match1) # for 1-1 pairs
```

The above p-value, which is > 0.05, tells us that we cannot reject the null hypothesis of the aline/non-aline groups having the same mortality rate. Assuming all assumptions of our modelling process are correct, we can infer from this that the use of an indwelling arterial catheter is not associated with a mortality benefit in these patients.
